%% ============================================================
%%  Chapter 3 — Methodology
%% ============================================================
\chapter{Methodology}
\label{ch:methodology}

This chapter describes the complete design of the HoneyIQ simulation
framework.
Section~\ref{sec:arch} presents the high-level architecture.
Section~\ref{sec:attacker} describes the attacker module.
Section~\ref{sec:threat} defines the composite threat-level metric and
reward function.
Section~\ref{sec:sedm} presents the Stage-Escalation Decision Matrix in
detail.
Section~\ref{sec:dqn_baseline} describes the DQN baseline.
Section~\ref{sec:env} documents the Gymnasium environment.
Section~\ref{sec:classifier} covers the Random Forest classifier.
Section~\ref{sec:eval_infra} describes the evaluation infrastructure.

%% ── 3.1 System Architecture ──────────────────────────────────
\section{System Architecture}
\label{sec:arch}

HoneyIQ is a closed-loop simulation in which an \emph{Attacker} and a
\emph{Defender} interact through a Gymnasium-compatible environment
(\texttt{CyberSecurityEnv}).
Figure~\ref{fig:architecture_text} shows the component layout and data flow.

\begin{figure}[ht]
  \centering
  \begin{tabular}{|p{0.85\textwidth}|}
    \hline
    \vspace{0.3cm}
    \textbf{CyberSecurityEnv} (gymnasium.Env wrapper)
    \begin{description}
      \item[Attacker:] TransitionModel (Markov chains) $\rightarrow$
            feature simulation $\rightarrow$
            $(a_t, k_t, \mathbf{x}_t)$
      \item[Threat:] compute\_threat\_level$(a_t, k_t, r_t, n_t) \rightarrow T \in [0,1]$
      \item[State:] $s_t \in \mathbb{R}^{24}$ (one-hot + scalars)
      \item[Defender:] AttackClassifier (RF) $+$ MatrixPolicy (SEDM) $\rightarrow$ action $\in \{0,\ldots,4\}$
      \item[Reward:] compute\_reward$(a_t, T, k_t, \text{attack\_type}) \rightarrow r_t$
    \end{description}
    \vspace{0.1cm}
    \textbf{MetricsCollector}: StepRecord $\rightarrow$ EpisodeRecord $\rightarrow$ CSV + plots
    \vspace{0.3cm} \\
    \hline
  \end{tabular}
  \caption{High-level architecture of the HoneyIQ simulation framework.}
  \label{fig:architecture_text}
\end{figure}

At each discrete time step $t$, the following sequence of operations is
executed:
%
\begin{enumerate}
  \item The attacker advances its Markov chains to produce a new
        $(a_t, k_t)$ pair and samples a 15-dimensional feature vector
        $\mathbf{x}_t$.
  \item The environment computes the threat level $T_t$ and constructs
        the 24-dimensional state vector $s_t$.
  \item The defender observes $s_t$ and selects a honeypot action.
  \item The reward $r_t$ is computed and returned to the defender.
  \item Metrics are recorded.
\end{enumerate}

This one-step interaction repeats for a configurable number of steps
(200 in evaluation, 500 in DQN training) before the episode terminates.

%% ── 3.2 Attacker Module ──────────────────────────────────────
\section{Attacker Module}
\label{sec:attacker}

\subsection{Attack Type and Kill Chain Enumerations}

The discrete state spaces of the attacker are defined by two \texttt{IntEnum}
classes:

\paragraph{AttackType.}
Ten categories inspired by the UNSW-NB15 classification taxonomy:
NORMAL (0), RECONNAISSANCE (1), ANALYSIS (2), FUZZERS (3), GENERIC (4),
EXPLOITS (5), SHELLCODE (6), BACKDOORS (7), DOS (8), WORMS (9).
Each category is assigned a severity weight $s_i \in [0, 1]$
(Table~\ref{tab:attack_severity_meth}).

\begin{table}[ht]
  \centering
  \caption{Attack types, severity weights, primary kill-chain stages,
           and kill-chain stage weights.}
  \label{tab:attack_severity_meth}
  \begin{tabular}{clccc}
    \toprule
    Index & Attack Type & Severity $s_i$ & Primary Stage & Stage Weight $w_k$ \\
    \midrule
    0 & NORMAL        & 0.00 & Reconnaissance (0) & 0.10 \\
    1 & RECONNAISSANCE & 0.20 & Reconnaissance (0) & 0.10 \\
    2 & ANALYSIS      & 0.25 & Weaponization (1)  & 0.20 \\
    3 & FUZZERS       & 0.35 & Delivery (2)       & 0.35 \\
    4 & GENERIC       & 0.40 & Delivery (2)       & 0.35 \\
    5 & EXPLOITS      & 0.70 & Exploitation (3)   & 0.55 \\
    6 & SHELLCODE     & 0.75 & Exploitation (3)   & 0.55 \\
    7 & BACKDOORS     & 0.80 & Installation (4)   & 0.70 \\
    8 & DOS           & 0.85 & Actions on Obj (6) & 1.00 \\
    9 & WORMS         & 0.90 & Cmd \& Ctrl (5)    & 0.85 \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{KillChainStage.}
Seven stages mapped to the Lockheed Martin Kill Chain:
RECONNAISSANCE (0), WEAPONIZATION (1), DELIVERY (2), EXPLOITATION (3),
INSTALLATION (4), COMMAND\_AND\_CTRL (5), ACTIONS\_ON\_OBJ (6).
Stage weights $w_k$ increase monotonically from 0.10 to 1.00, reflecting
the escalating danger of later kill-chain activity.

\subsection{Markov Transition Model}
\label{sec:transition}

Two row-stochastic base transition matrices are maintained:
$\mathbf{T}_A \in \mathbb{R}^{10 \times 10}$ (attack types) and
$\mathbf{T}_K \in \mathbb{R}^{7 \times 7}$ (kill chain stages).
At each step the attacker samples:
%
\begin{align}
  a_{t+1} &\sim \text{Categorical}\bigl(\mathbf{T}_A^{(\pi)}[a_t, \cdot]\bigr), \\
  \tilde{k}_{t+1} &\sim \text{Categorical}\bigl(\mathbf{T}_K^{(\pi)}[k_t, \cdot]\bigr).
\end{align}
%
A kill-chain floor constraint prevents unrealistic backward jumps:
%
\begin{equation}
  k_{t+1} = \max\!\bigl(\tilde{k}_{t+1},\;
                          \text{primary\_stage}(a_{t+1}) - 1\bigr).
  \label{eq:floor}
\end{equation}
%
This ensures that attack types associated with late kill chain stages cannot
appear at very early stages, maintaining temporal consistency.

\subsubsection{Intent Profiles}
\label{sec:intents}

Four intent profiles $\pi \in \{\text{STEALTHY}, \text{AGGRESSIVE},
\text{TARGETED}, \text{OPPORTUNISTIC}\}$ are encoded as element-wise
modifier matrices applied before row-renormalisation
(Table~\ref{tab:intents_meth}).

\begin{table}[ht]
  \centering
  \caption{Attacker intent profiles, behavioural characteristics, and
           primary attack preferences.}
  \label{tab:intents_meth}
  \begin{tabular}{lll}
    \toprule
    Intent & Attack Preference & Kill Chain Speed \\
    \midrule
    STEALTHY      & Recon, Analysis, Backdoors    & Slow; high self-loop probability \\
    AGGRESSIVE    & DoS, Worms, Exploits          & Fast; strong forward bias \\
    TARGETED      & Exploits $\to$ Shellcode $\to$ Backdoors & Direct exploitation chain \\
    OPPORTUNISTIC & Generic, Fuzzers (scattered)  & Moderate forward bias \\
    \bottomrule
  \end{tabular}
\end{table}

\textbf{STEALTHY}: the modifier matrix amplifies the self-loop probability
for RECONNAISSANCE and ANALYSIS, and increases the transition probability
from ANALYSIS to BACKDOORS.
Stage transitions are biased towards self-loops (the attacker dwells in each
stage longer), producing long-duration, low-severity campaigns.

\textbf{AGGRESSIVE}: forward transitions are amplified for DOS, WORMS, and
EXPLOITS.
Stage transitions are biased strongly forward, producing rapid escalation
through the kill chain.
The result is short, high-intensity campaigns that quickly reach ACTIONS\_ON\_OBJ.

\textbf{TARGETED}: transitions from EXPLOITS to SHELLCODE, SHELLCODE to
BACKDOORS, and BACKDOORS to WORMS are amplified, encoding a specific attack
chain.
The kill chain transitions are biased towards a fast path from EXPLOITATION
through INSTALLATION to COMMAND\_AND\_CTRL.

\textbf{OPPORTUNISTIC}: transitions to FUZZERS and GENERIC are amplified,
producing scattered attack-type sequences.
Stage transitions maintain moderate forward bias, resulting in a campaign that
explores multiple attack types without strong directionality.

\subsection{Network Feature Simulation}
\label{sec:features}

At each step the attacker samples a 15-dimensional feature vector
$\mathbf{x}_t \in \mathbb{R}^{15}$ from parametric distributions conditioned
on the current attack type.
The fifteen features are:
\texttt{dur}, \texttt{sbytes}, \texttt{dbytes}, \texttt{sttl},
\texttt{dttl}, \texttt{sloss}, \texttt{dloss}, \texttt{sload},
\texttt{dload}, \texttt{spkts}, \texttt{dpkts}, \texttt{swin},
\texttt{dwin}, \texttt{ct\_srv\_src}, \texttt{ct\_dst\_ltm}.

Selected distributions illustrate the per-attack signature design:

\begin{itemize}
  \item \textbf{RECONNAISSANCE}: \texttt{dur}~$\sim$~Exponential(rate=5),
        \texttt{ct\_dst\_ltm}~$\sim$~Poisson(30), reflecting rapid
        port-scanning activity.
  \item \textbf{DoS}: \texttt{sload}~$\sim$~Uniform(50\,000, 500\,000)\,bps,
        \texttt{dur}~$\sim$~Exponential(rate=10), capturing the high-bandwidth,
        short-burst pattern of flooding attacks.
  \item \textbf{BACKDOORS}: \texttt{dur}~$\sim$~Uniform(10, 3600)\,s,
        \texttt{sload}~$\sim$~Uniform(50, 500)\,bps, modelling long-lived,
        low-bandwidth covert channels.
  \item \textbf{WORMS}: \texttt{ct\_dst\_ltm}~$\sim$~Poisson(40),
        \texttt{spkts}~$\sim$~Poisson(200), reflecting spreading behaviour.
  \item \textbf{EXPLOITS}: \texttt{sbytes}~$\sim$~Uniform(1000, 50\,000),
        \texttt{sload}~$\sim$~Uniform(10\,000, 100\,000)\,bps, capturing
        the payload-heavy pattern of exploit delivery.
\end{itemize}

The 15 features are passed to the Random Forest classifier for attack-type
identification.
They are not directly included in the DQN state vector, which uses
ground-truth labels.

%% ── 3.3 Composite Threat Level and Reward ────────────────────
\section{Composite Threat Level and Reward Function}
\label{sec:threat}

\subsection{Composite Threat Level}
\label{sec:threat_level_def}

The threat level $T_t \in [0, 1]$ aggregates four signals into a single
normalised index:
%
\begin{equation}
  T_t = 0.45 \cdot s_{a_t}
      + 0.35 \cdot w_{k_t}
      + 0.15 \cdot r_t
      + 0.05 \cdot \min\!\left(1,\, \frac{n_t}{100}\right),
  \label{eq:threat}
\end{equation}
%
where:
\begin{itemize}
  \item $s_{a_t} \in [0, 0.90]$ is the attack-type severity weight
        (Table~\ref{tab:attack_severity_meth}).
  \item $w_{k_t} \in [0.10, 1.00]$ is the kill-chain stage weight.
  \item $r_t \in [0, 1]$ is the \emph{escalation rate}: the fraction of
        the last 20 steps that contained attack traffic.
        It captures the \emph{tempo} of the campaign independently of the
        specific attack types.
  \item $n_t$ is the cumulative attack count; the capped normalisation
        $\min(1, n_t / 100)$ provides a measure of cumulative campaign
        pressure that saturates after 100 attacks.
\end{itemize}

The weight assignment reflects a domain-informed prioritisation.
Attack severity (45\%) is the dominant signal because the attack type
directly indicates the attacker's capability and intent.
Kill chain stage (35\%) captures the attacker's progress through the
campaign, which determines the urgency of response.
Escalation rate (15\%) and cumulative count (5\%) provide temporal context
that distinguishes an isolated event from an ongoing campaign.

\paragraph{Threat bands.}
The threat level is partitioned into five bands used by the reward function:

\begin{center}
\begin{tabular}{lc}
  \toprule
  Band & Range \\
  \midrule
  Benign & $T < 0.15$ \\
  Low    & $0.15 \leq T < 0.35$ \\
  Medium & $0.35 \leq T < 0.55$ \\
  High   & $0.55 \leq T < 0.75$ \\
  Critical & $T \geq 0.75$ \\
  \bottomrule
\end{tabular}
\end{center}

\subsection{Reward Function}
\label{sec:reward}

The reward function $R(a, T, k, \text{attack\_type})$ encodes domain
knowledge as a structured signal with three layers.

\paragraph{Layer 1: Base matrix.}
The base reward is looked up from a $5 \times 5$ action-by-threat-band matrix
(Table~\ref{tab:reward_matrix_meth}).

\begin{table}[ht]
  \centering
  \caption{Base reward matrix (action $\times$ threat band).
           Values encode the desirability of each action--threat pairing.}
  \label{tab:reward_matrix_meth}
  \begin{tabular}{lccccc}
    \toprule
    Action & Benign & Low & Medium & High & Critical \\
    \midrule
    ALLOW & $+1.0$ & $+0.5$ & $-1.0$ & $-3.0$ & $-6.0$ \\
    LOG   & $+0.2$ & $+1.5$ & $+2.0$ & $+1.0$ & $-1.0$ \\
    TROLL & $-1.0$ & $+1.0$ & $+3.0$ & $+2.5$ & $+0.5$ \\
    BLOCK & $-2.0$ & $-0.5$ & $+1.5$ & $+3.5$ & $+5.0$ \\
    ALERT & $-3.0$ & $-1.0$ & $+0.5$ & $+2.0$ & $+6.0$ \\
    \bottomrule
  \end{tabular}
\end{table}

The matrix encodes graduated proportional responses: ALLOW is rewarded for
benign traffic but incurs increasing penalties as threat level rises.
ALERT receives the highest reward for critical threats ($+6.0$) and the
largest penalty for benign traffic ($-3.0$), reflecting its asymmetric cost
structure.
TROLL peaks at Medium threat (+3.0), incentivising engagement with
medium-severity attacks for intelligence gathering.

\paragraph{Layer 2: Late kill-chain amplifier.}
At kill chain stages Installation (4), Command \& Control (5), and
Actions on Objectives (6), negative rewards are amplified by a factor of
1.5$\times$:
%
\begin{equation}
  r \leftarrow 1.5 \cdot r \quad \text{if } r < 0 \text{ and } k \geq 4.
\end{equation}
%
This modifier encodes the escalating cost of defensive errors at advanced
campaign stages, where allowing or under-responding to an attack is
substantially more damaging than at early stages.

\paragraph{Layer 3: Attack-type bonuses.}
Specific action--attack-type combinations receive domain-knowledge bonuses:
%
\begin{itemize}
  \item TROLL on BACKDOORS, SHELLCODE, or WORMS: $+0.8$.
        These persistent attack types offer the highest intelligence value
        when the defender engages them for extended periods.
  \item BLOCK on WORMS: $+1.0$.
        Rapid containment of spreading malware is particularly valuable.
  \item LOG on RECONNAISSANCE: $+0.5$.
        Silent logging of reconnaissance provides intelligence without
        alerting the attacker.
  \item ALLOW on confirmed non-attack: $+0.5$.
        Correct passthrough of benign traffic is explicitly rewarded to
        discourage over-aggressive responses.
\end{itemize}

%% ── 3.4 Stage-Escalation Decision Matrix ─────────────────────
\section{Stage-Escalation Decision Matrix (SEDM)}
\label{sec:sedm}

The SEDM is a transparent, deterministic policy that maps the observable
environment state to a honeypot action through a five-step procedure.

\subsection{Design Rationale}

The SEDM is motivated by three observations:

\begin{enumerate}
  \item The Markov chain structure provides \emph{exact} escalation
        probabilities that encode the attacker's likely next move without
        requiring trajectory data.
  \item Kill chain stage is the single most informative variable for
        honeypot response: the appropriate action is qualitatively
        different at Reconnaissance (observe passively) versus
        Actions on Objectives (escalate immediately).
  \item Operational security practitioners require policies whose reasoning
        can be audited, documented, and modified by hand.
        A $7 \times 3$ matrix with three override rules satisfies this
        requirement; a 256-node neural network does not.
\end{enumerate}

\subsection{Algorithm}
\label{sec:sedm_alg}

\paragraph{Step 1 — Escalation risk.}
Query the intent-specific Markov transition model for the probability of
advancing to a strictly higher kill chain stage from the current stage $k$:
%
\begin{equation}
  \rho(k, \pi) = \sum_{k' > k} T^{(\pi)}_{kk'}.
  \label{eq:esc_risk_meth}
\end{equation}
%
This quantity is computed analytically from the stored transition matrix.
Under the STEALTHY intent, stage transitions are biased towards self-loops,
producing low escalation risk values.
Under the AGGRESSIVE intent, strong forward transitions produce high
escalation risk.

\paragraph{Step 2 — Band classification.}
Discretise $\rho$ into three bands:
%
\begin{equation}
  \text{band} =
  \begin{cases}
    0 \; (\text{Low})    & \rho < 0.35 \\
    1 \; (\text{Medium}) & 0.35 \leq \rho < 0.65 \\
    2 \; (\text{High})   & \rho \geq 0.65.
  \end{cases}
\end{equation}
%
The thresholds 0.35 and 0.65 partition the escalation risk into three
roughly equal tertiles.
The Low band covers stages with strong self-loops (primarily Reconnaissance
under Stealthy intent); High covers stages with strong forward momentum
(primarily Delivery through Installation under Aggressive intent).

\paragraph{Step 3 — Matrix lookup.}
Read the base action from the $7 \times 3$ SEDM
(Table~\ref{tab:sedm_matrix}):
%
\begin{equation}
  a_{\text{base}} = \text{SEDM}[k][\text{band}].
\end{equation}

\begin{table}[ht]
  \centering
  \caption{Stage-Escalation Decision Matrix.
           Rows are kill chain stages; columns are escalation risk bands.}
  \label{tab:sedm_matrix}
  \begin{tabular}{lccc}
    \toprule
    Stage & Low ($\rho < 0.35$) & Medium ($0.35 \leq \rho < 0.65$) & High ($\rho \geq 0.65$) \\
    \midrule
    RECONNAISSANCE   & ALLOW & LOG   & LOG   \\
    WEAPONIZATION    & LOG   & LOG   & TROLL \\
    DELIVERY         & LOG   & TROLL & TROLL \\
    EXPLOITATION     & TROLL & BLOCK & BLOCK \\
    INSTALLATION     & BLOCK & BLOCK & ALERT \\
    COMMAND\_\&\_CTRL & BLOCK & ALERT & ALERT \\
    ACTIONS\_ON\_OBJ & ALERT & ALERT & ALERT \\
    \bottomrule
  \end{tabular}
\end{table}

The matrix encodes three design principles:
%
\begin{itemize}
  \item \textbf{Proportionality}: responses escalate with both kill chain
        stage and escalation risk. No ALERT is issued before Exploitation
        even under high escalation risk; no ALLOW is issued after
        Weaponization.
  \item \textbf{Intelligence over containment at low stages}: LOG and TROLL
        are preferred over BLOCK at early stages to maximise intelligence
        before committing to containment.
  \item \textbf{Decisive containment at late stages}: ALERT is the
        dominant action from Installation onwards under medium or high
        escalation risk, reflecting the low tolerance for delay at advanced
        kill chain positions.
\end{itemize}

\paragraph{Step 4 — Override rules.}
Three override rules are applied sequentially after the matrix lookup:
%
\begin{description}
  \item[R1 (Normal traffic):]
        If \texttt{AttackType} = NORMAL, return ALLOW regardless of
        the matrix result.
        This ensures that confirmed benign traffic is never blocked.
  \item[R2 (High-impact attack types):]
        If \texttt{AttackType} $\in$ \{DOS, WORMS\}, upgrade the action
        by one level in the sequence
        ALLOW $\to$ LOG $\to$ TROLL $\to$ BLOCK $\to$ ALERT.
        These attack types cause rapid, widespread damage and warrant
        an immediately more aggressive response.
  \item[R3 (High attack frequency):]
        If the escalation rate $r_t > 0.80$, upgrade the action by one
        level.
        A sliding window showing that more than 80\% of recent steps
        involved attacks indicates a sustained, ongoing campaign that
        requires escalated containment.
\end{description}

Override rules R2 and R3 are applied conditionally: R2 is checked first,
and R3 is checked only if R2 was not triggered.
This precedence prevents double-upgrading within a single step.

\paragraph{Step 5 — Composite risk score.}
A composite risk score $\rho_c \in [0, 1]$ is computed for logging and
analysis.
It does not affect the action selection:
%
\begin{equation}
  \rho_c = 0.35 \cdot w_k
         + 0.35 \cdot \rho(k, \pi)
         + 0.15 \cdot s_a
         + 0.15 \cdot r_t.
  \label{eq:composite_risk}
\end{equation}

\subsection{Intent-Awareness}

The SEDM is intent-aware through the escalation risk computation
(Equation~\ref{eq:esc_risk_meth}).
The base matrix $\text{SEDM}[k][\text{band}]$ is identical across all
intents, but the band assigned to a given stage differs:
%
\begin{itemize}
  \item Under STEALTHY, stage transitions are slow.
        RECONNAISSANCE has low escalation risk ($\rho < 0.35$) → ALLOW.
        The defender correctly passes low-threat reconnaissance traffic.
  \item Under AGGRESSIVE, stage transitions are fast.
        Even RECONNAISSANCE may have medium escalation risk ($\rho \geq 0.35$)
        → LOG, reflecting the attacker's tendency to escalate quickly.
  \item Under TARGETED, the direct exploitation path produces consistently
        high escalation risk from EXPLOITATION onwards → BLOCK/ALERT
        dominates.
\end{itemize}

The escalation risk is the key mechanism through which the SEDM adapts to
different attacker behaviours without any learning or parameter updates.

\subsection{Worked Example}

Consider an observation with:
stage = EXPLOITATION (3), attack type = EXPLOITS, escalation rate = 0.65,
intent = AGGRESSIVE.

\begin{enumerate}
  \item \textbf{Escalation risk}: Under AGGRESSIVE intent,
        $T^{(\text{AGG})}_{3,4} + T^{(\text{AGG})}_{3,5} + T^{(\text{AGG})}_{3,6} = 0.72$
        (high forward bias in the transition matrix).
  \item \textbf{Band}: $\rho = 0.72 \geq 0.65 \Rightarrow$ band = High.
  \item \textbf{Matrix lookup}: SEDM[3][2] = BLOCK.
  \item \textbf{Override check}:
        EXPLOITS $\notin$ \{DOS, WORMS\} (R2 not triggered);
        escalation rate 0.65 $\leq$ 0.80 (R3 not triggered).
  \item \textbf{Final action}: BLOCK.
  \item \textbf{Composite risk}:
        $\rho_c = 0.35 \times 0.55 + 0.35 \times 0.72 + 0.15 \times 0.70 + 0.15 \times 0.65 = 0.654$.
\end{enumerate}

%% ── 3.5 DQN Baseline ─────────────────────────────────────────
\section{DQN Baseline}
\label{sec:dqn_baseline}

The DQN baseline provides a learned policy for comparison with the SEDM.
Its architecture and training hyperparameters are summarised in
Table~\ref{tab:hyperparams}.

\begin{table}[ht]
  \centering
  \caption{DQN hyperparameters used in all training experiments.}
  \label{tab:hyperparams}
  \begin{tabular}{ll}
    \toprule
    Parameter & Value \\
    \midrule
    State dimension & 24 \\
    Action space & 5 (\texttt{HoneypotAction}) \\
    Hidden layers & [256, 128, 64] \\
    Normalisation per layer & \texttt{LayerNorm} \\
    Activation & \texttt{ReLU} \\
    Weight initialisation & Kaiming uniform \\
    Replay buffer capacity & 15\,000 transitions \\
    Mini-batch size & 64 \\
    Discount factor $\gamma$ & 0.99 \\
    Learning rate & $10^{-3}$ (Adam) \\
    Target network update interval & 150 steps (hard copy) \\
    $\varepsilon$ initial / final / decay & 1.0 / 0.05 / 0.997 \\
    Gradient clip $\ell_2$ norm & 10.0 \\
    Loss function & Huber (SmoothL1) \\
    Training episodes & 300 \\
    Steps per episode & 500 \\
    Training intent & OPPORTUNISTIC \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Training Protocol}

Training proceeds for 300 episodes of 500 steps each, using the
OPPORTUNISTIC attacker intent.
OPPORTUNISTIC was chosen as the training distribution because its scattered
attack-type sequence provides broad coverage of the state space, encouraging
the policy network to learn a generalised Q-function rather than one
optimised for a specific escalation pattern.

At each step:
%
\begin{enumerate}
  \item The defender observes the 24-dimensional state vector.
  \item An action is selected via $\varepsilon$-greedy exploration.
  \item The transition $(s, a, r, s', \text{done})$ is pushed to the replay buffer.
  \item If the buffer contains at least 64 transitions, one gradient update
        is performed with a randomly sampled mini-batch.
  \item Epsilon is decayed: $\varepsilon \leftarrow \max(0.05, 0.997\varepsilon)$.
  \item Every 150 steps, the target network is hard-copied from the policy network.
\end{enumerate}

Metrics recorded per episode: total reward, detection rate, false-positive
rate, average threat level, average training loss.
Training metrics are saved to \texttt{logs/metrics.csv} and visualised as
six-panel training curves.

%% ── 3.6 Gymnasium Environment ────────────────────────────────
\section{Gymnasium Environment (\texttt{CyberSecurityEnv})}
\label{sec:env}

\texttt{CyberSecurityEnv} extends \texttt{gymnasium.Env}~\citep{towers2024gymnasium}
and implements the standard Gymnasium interface.

\subsection{Observation Space}

The observation space is $\text{Box}(24,)$ with \texttt{float32} dtype.
The 24-dimensional vector is constructed as follows:
%
\begin{equation}
  s_t = \underbrace{[\,e_{a_t}^{(10)}\,]}_{\text{attack type (one-hot)}}
       \;\oplus\;
       \underbrace{[\,e_{k_t}^{(7)}\,]}_{\text{stage (one-hot)}}
       \;\oplus\;
       \underbrace{\left[\,T_t,\;\tfrac{n_t}{100},\;r_t\right]}_{\text{scalars}}
       \;\oplus\;
       \underbrace{[\,e_{\pi}^{(4)}\,]}_{\text{intent (one-hot)}},
  \label{eq:state}
\end{equation}
%
where $e_x^{(d)}$ denotes a $d$-dimensional one-hot vector for category $x$.
The scalar features ($T_t$, $n_t/100$, $r_t$) lie in $[0, 1]$ by
construction.
Including the attacker intent in the state vector allows both the DQN and the
SEDM to condition their responses on the inferred attacker type, even though
the SEDM queries the Markov chain directly.

\subsection{Action Space}

The action space is \texttt{Discrete(5)}, one element per \texttt{HoneypotAction}.
The integer-to-action mapping is: 0\,=\,ALLOW, 1\,=\,LOG, 2\,=\,TROLL,
3\,=\,BLOCK, 4\,=\,ALERT.

\subsection{Episode Lifecycle}

\texttt{reset()} re-initialises the attacker Markov chains to their starting
states, clears the escalation history window, and resets the attack counter.
It returns the initial state vector and an \texttt{info} dictionary containing
the initial features and attack-type ground truth.

\texttt{step(action)} executes one simulation step:
%
\begin{enumerate}
  \item Calls \texttt{attacker.step()} to advance the Markov chains and
        sample a feature vector.
  \item Computes the threat level $T_t$ (Equation~\ref{eq:threat}).
  \item Computes the reward $r_t$ (Section~\ref{sec:reward}).
  \item Constructs the next state vector $s_{t+1}$
        (Equation~\ref{eq:state}).
  \item Increments the step counter; terminates the episode if
        \texttt{max\_steps} is reached.
  \item Returns \texttt{(next\_state, reward, terminated, truncated, info)}.
\end{enumerate}

Episodes terminate by truncation (\texttt{max\_steps} reached) rather than
by natural termination conditions.
The environment does not model attacker defeat or withdrawal, consistent with
the assumption that the attacker's Markov chain is independent of the
defender's responses.

%% ── 3.7 Attack Classifier ────────────────────────────────────
\section{Random Forest Attack Classifier}
\label{sec:classifier}

The \texttt{AttackClassifier} wraps a \texttt{RandomForestClassifier} from
scikit-learn with a \texttt{StandardScaler} preprocessing step.

\subsection{Training Data Generation}

Training data is generated entirely from the parametric feature distributions
(Section~\ref{sec:features}).
For each of the ten attack types, 600 samples are generated independently:
%
\begin{equation}
  \mathcal{D}_{\text{train}} = \bigcup_{a=0}^{9}
  \bigl\{(\mathbf{x}, a) \,:\, \mathbf{x} \sim p(\cdot \mid a)\bigr\}_{600},
\end{equation}
%
yielding a balanced dataset of 6\,000 samples.
\texttt{class\_weight=`balanced'} provides a further guard against any
residual class imbalance.
The evaluation set (2\,000 samples, 200 per class) is generated with a
different random seed (999) to ensure independence from training data.

\subsection{Role in the System}

In the current implementation, the classifier output is logged alongside
the ground-truth attack type at each step for comparison and monitoring.
The SEDM reads the attack type directly from the environment state vector
(which contains the ground-truth label).
Integrating classifier outputs into the decision pipeline is identified as
a key direction for future work (Section~\ref{sec:future}).

%% ── 3.8 Evaluation Infrastructure ───────────────────────────
\section{Evaluation Infrastructure}
\label{sec:eval_infra}

\subsection{Metrics}

Two granularity levels are maintained:

\paragraph{Step-level (\texttt{StepRecord}).}
A dataclass capturing per-step data:
\texttt{episode, step, action, reward, attack\_type, kill\_chain\_stage,
threat\_level, is\_attack, predicted\_attack, loss, escalation\_rate}.

\paragraph{Episode-level (\texttt{EpisodeRecord}).}
Aggregated per-episode statistics:
%
\begin{align}
  \text{detection\_rate} &= \frac{\text{TP}}{\text{TP} + \text{FN}}, \\
  \text{false\_positive\_rate} &= \frac{\text{FP}}{\text{FP} + \text{TN}},
\end{align}
%
where a True Positive (TP) is any step where the attack was detected
(action $\neq$ ALLOW) and the traffic was genuinely malicious.
A False Positive (FP) is any step where the defender responded with a
non-ALLOW action on benign traffic.

\subsection{Experimental Protocol}

\paragraph{SEDM evaluation.}
The SEDM is evaluated for 30 episodes per attacker intent (120 total),
each episode lasting 200 steps.
No training is required; the policy is fully specified by the
Table~\ref{tab:sedm_matrix} and the Markov chain transition model.
Results are aggregated per intent and reported as mean $\pm$ standard
deviation over the 30 episodes.

\paragraph{DQN training.}
The DQN agent is trained for 300 episodes of 500 steps each using the
OPPORTUNISTIC intent.
All hyperparameters are listed in Table~\ref{tab:hyperparams}.
Training metrics are logged to \texttt{logs/metrics.csv}.

\paragraph{Reproducibility.}
All random seeds are fixed: seed 42 for training data generation and
environment resets; seed 999 for classifier evaluation.
The NumPy and PyTorch random states are seeded at the start of each run.
