%% ============================================================
%%  Chapter 6 — Conclusion
%% ============================================================
\chapter{Conclusion}
\label{ch:conclusion}

\section{Summary of the Work}

This thesis presented HoneyIQ, a simulation framework for adaptive honeypot
management in which a structured, intent-driven attacker --- modelled on the
Lockheed Martin Cyber Kill Chain~\citep{hutchins2011intelligence} --- is
opposed by two defender policies: the Stage-Escalation Decision Matrix (SEDM)
and a Deep Q-Network (DQN) baseline.

The work was motivated by three observations.
First, contemporary cyber adversaries conduct structured, multi-stage campaigns
that progress predictably through reconnaissance, weaponization, delivery,
exploitation, installation, command-and-control, and actions on objectives.
Second, honeypots provide a uniquely valuable defensive layer that can detect,
observe, and deceive attackers without disrupting production systems, but their
effectiveness depends on the quality of the response policy.
Third, learned policies based on deep reinforcement learning --- while capable
of high performance --- are fundamentally opaque and therefore difficult to
deploy in operational security environments that require auditability and
explainability.

To address these challenges, this thesis made the following contributions.

\subsection{HoneyIQ Simulation Framework}

A fully self-contained, Gymnasium-compatible simulation environment implementing:
%
\begin{itemize}
  \item A 24-dimensional state space combining one-hot attack type, kill chain
        stage, and attacker intent with continuous threat level, attack count,
        and escalation rate.
  \item A five-action honeypot policy interface (ALLOW, LOG, TROLL, BLOCK,
        ALERT) mapping response severity from passive passthrough to immediate
        escalation.
  \item A domain-informed reward function with three layers: a $5 \times 5$
        base matrix, a late-stage amplifier for kill chain stages 4--6, and
        attack-type-specific bonuses encoding honeypot best practices.
  \item A comprehensive metrics layer computing per-step and per-episode
        statistics including detection rate, false-positive rate, and
        composite risk, with six visualisation types.
\end{itemize}

All experiments are fully reproducible without access to live network data.

\subsection{Intent-Aware Markov Attacker}

A coupled pair of Discrete-Time Markov Chains governing attack-type and
kill-chain-stage transitions, with four intent profiles encoded as element-wise
transition modifiers:
%
\begin{itemize}
  \item \textbf{Stealthy}: slow, low-severity campaigns biased toward
        reconnaissance and backdoor activity.
  \item \textbf{Aggressive}: fast, high-severity campaigns biased toward
        DoS, Worms, and Exploits.
  \item \textbf{Targeted}: focused exploitation chains progressing from
        Exploits through Shellcode to Backdoors.
  \item \textbf{Opportunistic}: scattered campaigns with moderate escalation
        speed and diverse attack types.
\end{itemize}

Network features are sampled from per-attack parametric distributions
calibrated to UNSW-NB15 signatures, providing a bridge between the simulation
and real-world traffic characteristics.
The kill-chain floor constraint (Equation~\ref{eq:floor}) ensures temporal
consistency by preventing unrealistic stage regression.

\subsection{Stage-Escalation Decision Matrix (SEDM)}

The principal technical contribution: an interpretable, deterministic, zero-
training-overhead policy implementing a five-step decision algorithm:
%
\begin{enumerate}
  \item Compute escalation risk analytically from the intent-specific Markov
        transition matrix.
  \item Classify the risk into Low, Medium, or High band.
  \item Perform a $7 \times 3$ matrix lookup (Table~\ref{tab:sedm_matrix}).
  \item Apply three prioritised override rules (R1: Normal $\to$ ALLOW;
        R2: DOS/WORMS $\to$ upgrade; R3: high rate $\to$ upgrade).
  \item Log a composite risk score for situational awareness.
\end{enumerate}

The SEDM's transparency is its defining property: every decision is traceable
to a specific matrix entry and override condition, enabling full audit without
post-hoc explanation approximations.

\subsection{Composite Threat-Level Metric}

A weighted combination (attack severity 45\%, kill chain stage 35\%,
escalation rate 15\%, cumulative count 5\%) providing a normalised threat
index that drives both the reward function and the SEDM's composite risk score.
The weight assignment reflects domain reasoning about the relative informational
value of each signal.

\subsection{DQN Baseline and Comparison}

A full DQN implementation (24 $\to$ 256 $\to$ 128 $\to$ 64 $\to$ 5 with
LayerNorm, experience replay, target network, Huber loss, and epsilon-greedy
exploration) trained for 300 episodes on the Opportunistic intent.
The DQN achieves comparable detection rates ($>$98.5\%) to the SEDM but requires
150\,000 training steps, produces an opaque policy, and exhibits undefined
false-positive rate characteristics due to the dense-attack training distribution.

\section{Key Findings}

\paragraph{Finding 1: Analytical escalation risk enables exact, training-free adaptation.}
The Markov chain structure provides exact, per-intent escalation probabilities
that the SEDM uses directly.
This avoids the data dependency of RL-based approaches while correctly
capturing the intent-specific differences in campaign tempo.
The result is a policy that adapts to four qualitatively different attack
profiles without any training or parameter tuning.

\paragraph{Finding 2: Detection rates above 99\% are achievable with a deterministic matrix.}
The SEDM achieves detection rates of 99.09--99.48\% across all four attacker
intents.
This performance is competitive with the DQN baseline and substantially
exceeds what a random policy would achieve.
The result challenges the common assumption that high detection rate requires
a learned, adaptive policy.

\paragraph{Finding 3: False positive rates reflect intent-specific traffic distributions.}
False positive rates vary from 3.33\% (Targeted) to 35.56\% (Stealthy),
reflecting the intrinsic difficulty of discriminating attack from borderline
benign traffic at early kill chain stages under stealthy intent profiles.
This variation is a property of the evaluation scenarios rather than a failure
of the SEDM: the same policy that produces few false alarms against
high-severity campaigns naturally produces more against stealthy campaigns
where the traffic boundary is less sharp.

\paragraph{Finding 4: Interpretability and detection performance are compatible.}
The SEDM achieves detection rates comparable to the DQN with full
interpretability, zero training overhead, and superior cross-intent
consistency.
This finding supports the recommendation of \citet{rudin2019stop} to prefer
interpretable models in high-stakes domains when the problem structure permits
a good interpretable solution --- and the kill chain model clearly provides
such a structure.

\paragraph{Finding 5: DQN specialises to the training distribution.}
The DQN's effective strategy of ``almost never ALLOW'' achieves high
detection on the Opportunistic training distribution but would not generalise
to scenarios with more balanced benign/attack traffic.
This specialisation is an artefact of the dense-attack training environment,
not an inherent limitation of DQN, but it illustrates the importance of
training distribution design for RL-based security policies.

\section{Implications for Practice}

The results of this thesis suggest several practical implications for
honeypot management system design:

\begin{enumerate}
  \item \textbf{Start with an interpretable policy.}
        The SEDM's performance, immediate operability, and auditability make
        it the appropriate default for initial deployment.
        A security team can review, understand, and modify the 7-row matrix
        without requiring machine learning expertise.

  \item \textbf{Use the kill chain model explicitly.}
        The strong correlation between kill chain stage and appropriate
        response severity (ALLOW at Reconnaissance, ALERT at C2 and beyond)
        is a robust design principle that should be reflected in any honeypot
        policy, whether interpretable or learned.

  \item \textbf{Calibrate expectations on false positives by attacker profile.}
        A single false positive rate threshold is not appropriate for all
        deployment contexts.
        Networks facing stealthy APT campaigns should tolerate higher false
        positive rates at early kill chain stages; networks facing opportunistic
        automated attackers can operate with much lower thresholds.

  \item \textbf{Use RL for refinement, not as a starting point.}
        A DQN trained alongside a deployed SEDM --- using live honeypot
        interactions as the data source --- could discover refinements to the
        matrix that are not immediately obvious to human designers.
        This staged approach separates the need for immediate interpretability
        (SEDM) from the potential for long-term optimisation (DQN).
\end{enumerate}

\section{Closing Remarks}

HoneyIQ demonstrates that the mathematical structure of the Markov attacker
model --- specifically, the analytically computable escalation probabilities
derived from intent-specific transition matrices --- provides sufficient
information to design a high-quality, interpretable honeypot policy without
any machine learning.
The Stage-Escalation Decision Matrix translates the probabilistic dynamics
of the kill chain into a human-readable response protocol that aligns with
honeypot best practices: engaging low-severity traffic passively, tarpitting
persistent attackers for intelligence, and escalating decisively as campaigns
reach critical kill-chain stages.

The simulation infrastructure provides a principled, self-contained
foundation for future research.
Replacing synthetic feature distributions with real data fits, integrating
the classifier output into the decision pipeline, modelling an adaptive
attacker who responds to the defender's actions, and extending to a multi-node
network topology are natural next steps that would progressively close the gap
between simulation and operational deployment.

As networked systems face increasingly sophisticated adversaries, the
combination of structured attacker models, analytically-grounded decision
policies, and rigorous simulation-based evaluation offers a principled path
toward honeypot management that is simultaneously effective, trustworthy,
and deployable in the operational environments that need it most.
